# Hello bookdown 

All chapters start with a first-level heading followed by your chapter title, like the line above. There should be only one first-level heading (`#`) per .Rmd file.

## Setup

Make sure you tell GitHub that the web site is not to be build via Jekyll,
since the **bookdown** HTML output is already a standalone website. See
section 6.3 of [bookdown](https://bookdown.org/yihui/bookdown/github.html) for
details.

```{r}
# create a hidden file .nojekyll
# to tell GitHub that the website is not to be build via Jekyll
a_file <- file.path(getwd(), ".nojekyll")
if (!file.exists(a_file)) file.create(a_file)
```

## A section

All chapter sections start with a second-level (`##`) or higher heading followed by your section title, like the sections above and below here. You can have as many as you want within a chapter.

### An unnumbered section {-}

Chapters and sections are numbered by default. To un-number a heading, add a `{.unnumbered}` or the shorter `{-}` at the end of the heading, like in this section.

```{r}

```

