# (PART) Part I {-}

# Introduction {#intro}

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2, quietly = TRUE)
library(ggdag, quietly = TRUE)
```


## A Brief History


## Data Examples


### Mortality Rates by Country


### National Center for Education Statistics


### Reducing Alcohol Consumption


#### The What-If? Study


##### The Double What-If? Study


The code can be found in the file `doublewhatifsim.R` and is verified
against the author's file just below.

```{r include=FALSE}
a_dir <- file.path(dirname(getwd()), "FundamentalsCausalInference_docs",
                   "Brumback FOCI Website Material", "Chapter 1")
load(file.path(a_dir, "doublewhatifdat.RData"))
```


The DAG for the *Double What-If?* study in the `dagitty` version is

```{r}
scm <- list()
scm <- within(scm, {
  the_nodes <- c("U" = "Unmeasured, healthy behavior (U=1)", 
                 "AD0" = "Adherence time 0", 
                 "VL0" = "Viral Load time 0", 
                 "T" = "Naltrexone (T=1)", 
                 "A" = "Reduced drinking (A=1)", 
                 "AD1" = "Adherence time 1", 
                 "VL1" = "Viral Load time 1")
  coords <- data.frame(
    name = names(the_nodes),
    x = c(2, 3, 4, 1, 2, 3, 4),
    y = c(2, 2, 2, 1, 1, 1, 1)
  )
  dag <- dagify(
    AD0 ~ U,
    VL0 ~ AD0,
    A ~ `T` + U,
    AD1 ~ A,
    VL1 ~ AD0 + AD1 + U,
  outcome = "VL1",
  exposure = "T",
  latent = "U",
  coords = coords,
  labels = the_nodes)
  
  # status' colors
  colrs <- c("latent" = "palevioletred", "exposure" = "mediumspringgreen", 
             "outcome" = "cornflowerblue")
  # plot the DAG
  plot <- dag %>% 
    tidy_dagitty() %>%
    ggdag_status(color = status, text = TRUE,  node_size = 10, text_size = 3, 
                 text_col = "black") +
    scale_color_manual(values = colrs, na.value = "honeydew3") +
    scale_fill_manual(values = colrs, na.value = "honeydew3") +
    ggdag::theme_dag_blank(panel.background = 
                             element_rect(fill="snow", color="snow")) +
    theme(legend.position = "bottom",
          legend.title = element_blank()) +
    labs(title = "The Double What-If? Study")
})
scm$plot
```

and the code for `doublewhatifsim.R` is

```{r file="lib\\fci_01-A_doublewhatifsim.R"}

```


which we verify against the raw data set provided by the author.

```{r}
bb <- doublewhatifdat
d <- doublewhatifsim()
stopifnot(identical(d, bb))
```


### General Social Survey


### A Cancer Clinical Trial


## Exercises

Please note that the solutions to exercises of @brumback2022 below are my own 
and have not been verified or approved in any way by the author.

### Exercise 1

8 Experiment

### Exercise 2

9 Analogy

### Exercise 3

2 Consistency: *They are well known to cause substantial weight gain*. That is,
observed by different persons, different places and time.

6 Plausability

7 Coherence

### Exercise 4

8 Experiment

9 Analogy

### Exercise 5

3 Specificity

### Exercise 6

5 Biological gradient: *increasing levels of physical activity are more strongly
protective*

8 Experiment

### Exercise 7

6 Plausability

4 Temporality

### Exercise 8

4 Temporality
