# Backdoor Method via Standardization {#backdoor}


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(dagitty)
library(ggplot2)
library(ggdag)
library(gt)
```



```{r include=FALSE}
# the directory of documentation for chapter 1
dir_docs01 <- file.path(dirname(getwd()), "FundamentalsCausalInference_docs",
                   "Brumback FOCI Website Material", "Chapter 1")
# directory of data files
dir_data <- file.path(getwd(), "data")
# directory for functions
dir_lib <- file.path(getwd(), "lib")
```


```{r message=FALSE, warning=FALSE, include=FALSE}
source(file = file.path(dir_lib, "boot_utils.R"), 
       local = knitr::knit_global())
source(file = file.path(dir_lib, "fci_06-A_bootstand.R"), 
       local = knitr::knit_global())
source(file = file.path(dir_lib, "fci_01-A_doublewhatifsim.R"), 
       local = knitr::knit_global())
source(file = file.path(dir_lib, "gt_measures.R"), 
       local = knitr::knit_global())
load(file.path(dir_docs01, "whatifdat.RData"))
```


## Standardization via Outome Modeling

> Standardization vis via outcome modelingis one way to estimate $E(Y(t))$


$$
\begin{align*}
&\text{by double expectation theorem} \\
E(Y(t)) &= E_H E(Y(t) \mid H) \\
&\text{by independence of T given H, (6.1)} \\
&= E_HE(Y(t) \mid T=t, H) \\
&\text{by consistency assumption} \\
&= E_HE(Y \mid T=t, H)
\end{align*}
$$

and with a binary dataset we can write

$$
\begin{align*}
E_H E(Y \mid T=t, H) = E(Y \mid T=t, H = 0) P(H = 0) + E(Y \mid T=t, H = 1) P(H = 1)
\end{align*}
$$


We use `R` to compute the standardized estimates with the following script


```{r eval=TRUE, echo=TRUE, file="lib\\fci_06-A_bootstand.R"}

```


**Example: What-if? Study**

and for the What-If? study this gives


```{r}
whatif.out <- bootstand(whatifdat, formula = Y ~ A + H + A*H)
```

and we compare with the author's

```{r}
d <- data.frame(
  name = c("EY0", "EY1", "RD", "RR"),
  auth = c(0.375, 0.289, -0.086, 0.77),
  est = whatif.out$est
)
stopifnot(sum(abs(d$auth - d$est)) < 0.01)
```

and the results are presented in table 6.1

```{r}
gt_measures(whatif.out, 
            title = "Table 6.1", 
            subtitle = "What-If Study: Standardized Estimates")
```

where we observe a reduction of the viral load but the difference is not
statistically significant.


**Example: Double What-if? Study**


```{r}
dataDoubleWhatIf <- doublewhatifsim()
```


```{r}
doublewhatif.out <- bootstand(dataDoubleWhatIf, formula = VL1 ~ A + AD0 + A*AD0)
```


```{r}
gt_measures(doublewhatif.out, 
            title = "Table 6.2", 
            subtitle = "What-If Study: Standardized Estimates")
```



> For comparisons, we repeat the standardization with $H = VL_0$



```{r}
doublewhatif.out <- bootstand(dataDoubleWhatIf, formula = VL1 ~ A + VL0 + A*VL0)
```


```{r}
gt_measures(doublewhatif.out, 
            title = "Table 6.3", 
            subtitle = "What-If Study: Standardized Estimates")
```


### Average Effect of Treatment on the Treated
