# Functions {#functions}

All the scripts for the functions below can be found at 
[study project](https://github.com/FrankLef/FundamentalsCausalInference), mostly in the
`/lib` folder.

The functions are shown in order of appearance in the chapter where they are
first used.

## Chapter 2

The file `boot_utils` includes several utility functions used by the bootstrapping
functions.


```{r eval=FALSE, file="lib\\boot_utils.R"}

```


The file `gt_utils` includes utility functions to format tables.


```{r eval=FALSE, file="lib\\gt_utils.R"}

```




## Chapter 3

### `gt_measures.R`

```{r eval=FALSE, file="lib\\gt_measures.R"}

```


## Chapter 4

### `gt_measures_colgrp.R`


```{r eval=FALSE, file="lib\\gt_measures_colgrp.R"}

```


### `ggp_measures_modif.R`

```{r eval=FALSE, file="lib\\ggp_measures_modif.R"}

```


### `mc_beta_effect_measures.R`


```{r eval=FALSE, file="lib\\fci_04-B_mc_beta_effect_measures.R"}

```


### `ggp_venn_sim.R`


```{r eval=FALSE, file="lib\\ggp_venn_sim.R"}

```


### `ggp_betasim.R`


```{r eval=FALSE, file="lib\\ggp_betasim.R"}

```


## Chapter 5

### `ggp_dag.R`


```{r eval=FALSE, file="lib\\ggp_dag.R"}

```

### `gt_probs.R`


```{r eval=FALSE, file="lib\\gt_probs.R"}

```


## Chapter 6

For the `simdr` and `mc_standdr`, see the appendix on the doubly robust 
standardization simulation at [mc_standdr](#mc_standdr) for scripts and 
explanations.


## Chapter 7


```{r eval=FALSE, file="lib\\gt_measures_rowgrp.R"}

```
